Within and beyond: chert procurement patterns during the Upper Palaeolithic in Southwesternmost Iberia

### 2024-11-04 analysis/paper/paper.qmd
Reviews from reviewer #1

S_layerdata <- S_layerdata %>% 
  mutate_at("Date", str_replace, "20", "") %>%
  dplyr::rename("YEAR" = "Date") %>% 
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
  select(ID, Layer) %>% 
  distinct(ID, .keep_all = TRUE)

@@

shelter_field_data <- left_join(S_field_data, S_layerdata, by = "ID")
S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID")

@@

The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated to c. 24-22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017].

@@

The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings. The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b). These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules. Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.

@@

Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes (Bicho et al., 2013). The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.

@@

To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas. These two areas were chosen since both have detailed spatial information, well-defined chronological sequences, and previous studies providing lithic technology, faunal, and geoarchaeological data.

@@

Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~35-50% of the lithic assemblage from the Solutrean layers A, B, and C. Based on previous lithic studies in the Terrace and Shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].

Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs. Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020]. Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].

@@

S_comp <- S_comp %>% 
 mutate(Rawmaterial = case_when(
    RMTYPE == "Other" ~ "Other",
    RMTYPE == "TYPE 1" ~ "Chalcedony",
    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
  dplyr::mutate(Layer = case_when(
    Layer == "c" ~ "C",
    TRUE ~ Layer)) %>% 
  filter(!Layer %in% c("D", "Z", "No info")) # Choosing only Solutrean occupation layers (A, B and C)

@@

Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=5318). The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins. Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrino_materias_2013]. Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data. The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent geological survey efforts [@belmiroCreatingFramesReference2023]. Hence, an extensive macroscopic and petrographic collection was already available for comparative analysis with archaeological materials, employing consistent methodologies. This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.

@@

The macroscopic analysis of the archaeological materials was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create types and subtypes based on similar macroscopic characteristics which have been used in previous studies to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1). In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations. This first step was applied to a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 8 to 4) and 1665 chert artefacts from the Solutrean sequence of the Shelter area (layers C to A).

While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversitaDevelopmentMultiparametricCharacterisation2019]. Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016]. For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), they were analysed using the individual sample dataset, but grouped in a category of indeterminate cherts (INDET).

As seen in @tbl-indet-freq, level 4 shows the higher percentages of indeterminate artefacts, corresponding to \~40% of the total sample (n=1027). This is due to the frequent alterations of the artifacts, especially related to patinas and alteration rinds, pits, and fire/heat alterations. For all other levels in the Terrace, indeterminate represents on average \~28% of the chert artefacts, while in the Shelter the frequency of INDET is slightly lower (\~20-25%).

@@

Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony types, present throughout the stratigraphy in the Terrace and Shelter areas (a summarised characterization table can be found in the Supplementary Information, Online Resource 3). A distinct type was identified exclusively in the Shelter area, with only a limited presence in the Terrace (n=2), and 16 varieties, each consisting of one to three samples, were identified across various levels of both the Terrace and Shelter. The latter were clustered into a category named Trace Lithotypes (TL). Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts that are likely not local due to significant differences in appearance, fossil content, and petrography when compared to the cherts available in the geological setting located within a 30 km radius of Vale Boi.

### Cherts congruent with local sources

The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (@fig-local-types). The chalcedony (Type 1; @fig-local-types a) is identified as a massive, fibrous variety without allochems (@fig-local-types a2). It represents 6.6% of the overall Terrace cherts (excluding INDET) and 2.6% in the Shelter. It presents a colourless and translucent appearance, frequently exhibiting partial or complete white patination (@fig-local-types a1). This type is marked by extensive crystallizations, irregularities, and fractures, coupled with post-depositional alterations that challenge the identification and preservation of the cortex. Both macroscopic and petrographic analyses confirm its congruence with hydrothermal chalcedony located within lenses in Triassic volcanic formations, which are similarly marked by irregularities and fractures.

```{r}
#| label: tbl-cortex_tab
#| tbl-cap: Frequency of cortex type and thickness by raw material from the Terrace area.

@@

The cherts are massive micro-cryptocrystalline quartz from a marine environment (@fig-local-types b-h). They represent \~60% of the overall Terrace cherts and \~80% of the chert in the Shelter. Macroscopically, these types show variability which resulted in their separation in four different macroscopic types (Type 2 to 5). In general, these are opaque to sub-translucent cherts, with colours varying from brown, yellow, pale red, and grey, with the frequent presence of spicules and iron oxides. The differences are in their colours, nodule morphology, frequency, and variety of bioclasts (Type 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiroCreatingFramesReference2023]. Whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source (@tbl-cortex_tab). Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within Type 5. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Type 5 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from Type 5. Porosity varies between 10% and 1%, of vug type. Despite the petrographic homogeneity between local types, this macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.

Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic (Carixian) and Middle Jurassic Malhão formations of the Algarve basin. Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils. The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrops (@fig-map b). As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape. This is especially relevant since the inland outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].

```{r}
#| label: fig-local-types
#| fig-cap: Variability of local chert types and main petrographic characteristics. a) Type 1 macroscopic appearance (a1) and thin section (a2). b) Type 3 macroscopic appearance (b1) and thin section (b2). c-d) Type 2 macroscopic appearance (c1-d1) and thin section (c2-d2). e-f) Type 4 macroscopic appearance (e-f1) and thin section (f2-f3). g-h) Type 5 macroscopic appearance (g-h1) and thin section (h2-h3).

@@

### Cherts not congruent with local sources

@@

```{r}
#| label: fig-distance-types
#| fig-cap: Variability of non-local types and main petrographic characteristics. a-b) Type 6 macroscopic appearance (a1-b1) and thin section (a2-b2); b1-b2 correspond to a sample covered 100% in white patina. c-d) Type 7 macroscopic appearance (c1-d1) and thin section (c2-d2) of peloidal facies (c1-c2) and oolitic facies (d1-d2). e-f) Type 8 macroscopic appearance (e-f1) and thin section (f2-f3); white and black arrows correspond to foraminifera. g) Type 9 macroscopic appearance (g1) and thin section (g2). Type 10 macroscopic appearance (h1) and thin section (h2).

@@

Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022]. These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit. Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%). From these cherts, only the dark grey nodules (RT231) showed identifiable fossils which were rare ostracods.

@@

**Online resource 4**: Statistical analysis of Terrace chert types.

>>>>

S_layerdata <- S_layerdata %>% 
  mutate_at("Date", str_replace, "20", "") %>%
  dplyr::rename("YEAR" = "Date") %>% 
  unite(col='ID', cols_to_concat_shelter, sep="-", remove=FALSE) %>% 
  select(ID, Layer) %>% 
  distinct(ID, .keep_all = TRUE) %>%
  filter(!Layer %in% c("D", "Z"))

@@

shelter_field_data <- left_join(S_field_data, S_layerdata, by = "ID")
S_dataset <- left_join(S_RM_data, shelter_field_data, by = "ID")
S_dataset <- S_dataset %>% 
  mutate_at(vars(Layer), ~replace_na(., "ND")) %>% 
  filter(!Layer == "ND")

@@

The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the most intensive occupation, with three, well-preserved, archaeological horizons (layers C to A) and dated to c. 24-22 ka cal BP [@cascalheiraValeBoiAlgarve2013; @cascalheiraTecnologiaLiticaSolutrense2010; @cascalheira_2017]. These occupation levels were identified under blocks of limestone, which collapsed from the rockshelter ceiling [@cascalheiraTecnologiaLiticaSolutrense2010].

Both these areas have been previously interpreted as seasonal residential camps, repeatedly used for extended stays, due to the abundance of lithic debitage, stone tools, heat-cracked rocks related to grease rendering activities, large quantities of faunal remains (both marine and terrestrial) and the presence of ornaments and portable art [@manneIntensiveSubsistencePractices2012]. The analysis of lithic assemblages and retouched frequency of lithic assemblages from the Terrace and Shelter areas also corroborate this interpretation [@cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009; @cascalheira_2017]. Cascalheira et al. [-@cascalheira_2017] show that, with the exception of the Early Gravettian occupations of the Terrace area and Magdalenian occupation of the Shelter, all other UP occupations are composed of high-density assemblages with low degree of retouch, correspondent to an residential base-camp occupation, for extended periods of time.

@@

The availability of regional resources containing cherts and the characterization of their lithologies has already been published by our team [@belmiroCreatingFramesReference2023], and we herein present the summarized main findings. The results showed the existence of abundant chert nodules and beds found in the Lower Jurassic (Carixian) formation and Upper Jurassic (Kimmeridgian) formation, mostly outcropping in the cliff and beach areas of Westernmost Algarve, at \~20 km southwest from the site of Vale Boi, and with one single inland outcrop located \~8 km east from the site (@fig-map b). The Lower Jurassic nodules and beds can be found in abundant outcrops, but also frequently in sub-primary or secondary deposition settings in proximity to the outcrops possibly due to natural erosion. However, in the case of the inland outcrop of Ferrel, the outcrops is partially destroyed due to recent human action, and chert can be found scattered as chunks or larger nodules. The Upper Jurassic nodules can be found sub-primary deposition and of easy access, or in secondary deposition. These cherts are frequently opaque with yellow, grey, purple, or red colours, occasionally with bands or laminations, and with frequently visible sponge spicules. Although abundant, these chert nodules are frequently small (\~4-10 cm in length), with the largest nodules being \~20 cm, and frequently characterized by high levels of dolomitization and fractures.

@@

Chalcedony outcrops \~10 km west of Vale Boi (@fig-map b), within the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013], although previous studies also note the presence of chalcedony (with frequent presence of cortex) in the immediate vicinity of the site [@pereira_territory_2016-1; @gibajabaoProvenienceTechnologyMorphology2013]. The chalcedony frequently has fractures and fissures and is characterized by a fibrous quartz structure, without inclusions or fossils.

@@

To characterize the siliceous raw materials used during the UP occupations at Vale Boi and to identify their geological sources, we focused on chalcedony and chert archaeological materials from the Terrace and Shelter areas. These two areas were chosen since both have detailed spatial information, well-defined chronological sequences, and previous studies providing lithic technology, faunal, and geoarchaeological data. For this, we focused on layers that spanned the UP occupations, characterized by good preservation, a significant number of lithic finds and with radiocarbon dating (as mentioned in section Site description).

@@

Using the field database with all individually coordinated lithic artefacts (all pieces larger than \~2 cm; [@belmiroProtoSolutreanLithicTechnology2020]), we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels of the Terrace area (@fig-rm-comparison a), chert represents only \~15-20% of the lithic assemblage, while in the Shelter area (@fig-rm-comparison b), chert represents \~35-50% of the lithic assemblage from the Solutrean layers A, B, and C. Based on previous lithic studies in the Terrace and Shelter areas, the remaining 80% of other raw materials are mostly composed of quartz and greywacke [@belmiro_gravettian-solutrean_2021; @cascalheiraTecnologiaLiticaSolutrense2010; @marreirosPrimeirasComunidadesHomem2009].

Although systematic raw material research has not been conducted on these materials, according to the regional geological context, they have been interpreted as local and readily accessible in the vicinity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and unknapped chunks or slabs, and have been interpreted in great part as functionally specialised in activities related to the fragmentation of bones in the case of quartz and greywack [@cascalheira_2017], or in grease rendering in the case of quartz [@manneIntensiveSubsistencePractices2012]. Previous studies from the Terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiroProtoSolutreanLithicTechnology2020]. Based on this, despite the smaller amounts of individually plotted chert artefacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site, such as a Vale Comprido point, the fossil-director for the Proto-Solutrean in Western Portugal [@belmiro_gravettian-solutrean_2021].

@@

S_comp <- S_comp %>% 
 mutate(Rawmaterial = case_when(
    RMTYPE == "Other" ~ "Other",
    RMTYPE == "TYPE 1" ~ "Chalcedony",
    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert")) %>% # Collapsing all cherts into one single category
  dplyr::mutate(Layer = case_when(
    Layer == "c" ~ "C",
    TRUE ~ Layer)) %>% 
  filter(Layer %in% c("A", "B", "C")) # Choosing only Solutrean occupation layers (A, B and C)

@@

Macroscopic analysis, while revealing several limitations related to the absence of quantitative variables [@bustilloMacroscopicClassificationFlint2009], proved to be advantageous given the extensive sample size of chalcedony and chert lithics under examination (n=4458). The distinct features of various local cherts facilitate their differentiation from cherts of neighbouring origins. Consequently, a macroscopic approach offered a non-destructive, efficient, and cost-effective method for characterizing all materials [@bustilloMacroscopicClassificationFlint2009; @tarrino_materias_2013]. Therefore, we implemented an initial phase of macroscopic analysis, subsequently augmented by petrographic study to diminish the subjectivity of the analysis and enhance the material characterization with petrographic data. The synergistic application of these two methods has been previously validated in the characterization of Algarve region cherts, following recent geological survey efforts [@belmiroCreatingFramesReference2023]. Hence, an extensive macroscopic and petrographic collection was already available for comparative analysis with archaeological materials, employing consistent methodologies. This reference collection is the Lusolit lithotheque, hosted physically at ICArEHB (University of Algarve, Faro) and digitally at www.lusolit.icarehb.com, and aided in the characterization and source attribution of cherts.

@@

The macroscopic analysis of the archaeological materials was divided into two steps. The first step included a preliminary characterization using a hand lens of 10x magnification. This step allowed us to create types and subtypes based on similar macroscopic characteristics frequently acknowledged to be useful to describe and discern between different types of chert; these included characteristics such as colour, translucency, and feel [@luedtke_archaeologists_1992]-the complete data dictionary for the recorded variables can be found in the Supplementary Information (Online Resource 1). In the second step, we used a Nikon SMZ25 stereomicroscope to observe each sample in further detail, which allowed us to better characterize the artefacts (especially regarding inclusions and alterations), and the previously established types - the data dictionary used for the individual samples can also be found in the Supplementary Information (Online Resource 1). Given the inherent macroscopic variability of the geological samples and archaeological specimens, as well as the number of samples, the database used for the individual analysis was simplified and focused on collecting data related to weight, cortex, and alterations. This first step was applied to a total of 4458 artefacts: a sample of 3627 chert artefacts from the UP sequence of the Terrace area (levels 7 to 4) and 831 chert artefacts from the Solutrean sequence of the Shelter area (layers A, B, C; all samples without an attributed layer on the database were excluded from the analysis).

While the advantages of employing a macroscopic methodology for analysing the cherts of Vale Boi are notable, a limitation of this approach is its inability to unequivocally characterize and categorize artefacts that exhibit surface alterations typical of an open-air site, including those related to fire, since weathered and altered surface may destroy visible structures and even alter the chemical composition [@delluniversitaDevelopmentMultiparametricCharacterisation2019]. Other studies frequently exclude samples that display intense fire damage or several surface alterations [@delluniversitaDevelopmentMultiparametricCharacterisation2019; @gomezdesolerPanadellaChertMontmaneu2020; @sotoProcurementMobilityLate2016]. For the present study, whenever artefacts displayed extensive alterations and could not be unequivocally characterized (@fig-terrace-alterations), and given the unexistence of a reference collection for local altered cherts, they were analysed using the individual sample dataset (Online Resource 1) but grouped in a category of indeterminate cherts (INDET). This category included key alterations such as heat/fire related alterations which clearly impacted the macroscopic characteristics of the samples, fully changing its color, luster and texture, and covering them in crazings (e.g., @fig-terrace-alterations a). Similarly, it also included alteration rinds that extended through most of the sample, impeding the identification of its original color, texture, translucency, inclusions and fossils (e.g., @fig-terrace-alterations b) and the presence of extended pitting. However, whenever surface alterations were present in only a limited area of the sample leaving recognizable macroscopic features and fossils, or when key elements were distinguishable to make a characterization (e.g., reddening in the Type 2 cherts, Online Resource 2, figure s5), the samples were not collapsed into the INDET category.

As seen in @tbl-indet-freq, level 4 shows the highest percentages of indeterminate artefacts, corresponding to \~40% of the total sample (n=1037). This is due to the frequent alterations of the artifacts, especially related to patinas and alteration rinds, pits, and fire/heat alterations. For all other levels in the Terrace, indeterminate represents on average \~28% of the chert artefacts, while in the Shelter the frequency of INDET is slightly lower (\~20-25%).

@@

Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony types, present throughout the stratigraphy in the Terrace and Shelter areas (a summarised characterization table can be found in the Supplementary Information, Online Resource 3). A distinct type (Type 11) was identified exclusively in the Shelter area, with only a limited presence in the Terrace (n=2), and 16 varieties, each consisting of one to three samples, were identified across various levels of both the Terrace and Shelter. The latter were clustered into a category named Trace Lithotypes (TL). Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts that are likely not local due to significant differences in appearance, fossil content, and petrography when compared to the cherts available in the geological setting located within a 30 km radius of Vale Boi. @tbl-type-source shows the different types of identified cherts and possible sources whenever attribution was possible.

```{r}

#| label: tbl-type-source
#| tbl-cap: Chert types identified during the analysis, and potential sources. The distance column refers to the distance of the potential source from Vale Boi, as the crow flies. Only the TL samples with possible identified sources were included in the table.

tibble::tribble(
  ~"Chert type", ~"Potential source", ~"Distance from source",
  "Type 1", "Triassic volcanic formations ", " ",
  "Type 2-5", "Lower Carixian formations, Southwestern Portugal", "<20 km",
  "Type 6", "Upper Cretaceous formations, Central Portugal", "~215-250 km",
  "Type 7", "Betic Systems, Southern Spain", "~250 km",
  "Type 8", "-", "-",
  "Type 9", "Betic Systems, Southern Spain", "~250 km",
  "Type 10", "-", "-",
  "Type 11", "-", "-",
  "TL01", "South Portuguese Zone, Southern/Central Portugal", ">100 km",
  "TL10-11", "Upper Cretaceous formations, Central Portugal", "~215-250 km",
  "TL15-16", "Betic Systems, Southern Spain", "~250 km"
) |>
  knitr::kable()

### Cherts congruent with local sources (Types 1-5)

The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (@fig-local-types). The chalcedony (Type 1; @fig-local-types a) is identified as a massive, fibrous variety without allochems (@fig-local-types a2). It represents 6.6% of the overall Terrace cherts (excluding INDET) and 2.6% in the Shelter. It presents a colourless and translucent appearance, frequently exhibiting partial or complete white patination (@fig-local-types a1). This type is marked by extensive crystallizations, irregularities, and fractures, coupled with post-depositional alterations that challenge the identification and preservation of the cortex. Both macroscopic and petrographic analyses confirm its congruence with hydrothermal chalcedony located within lenses in Triassic volcanic formations, in the known outcrops of Monte do Tolo, Taboal and Cerro Velho (\<10 km west of Vale Boi, as the crow flies; @fig-map b), which are similarly marked by irregularities and fractures.

The cherts are massive micro-cryptocrystalline quartz from a marine environment (@fig-local-types b-h). They represent \~60% of the overall Terrace cherts and \~80% of the chert in the Shelter. Macroscopically, these types show variability which resulted in their separation in four different macroscopic types (Type 2 to 5). In general, these are opaque to sub-translucent cherts, with colours varying from brown, yellow, pale red, and grey, with the frequent presence of spicules and iron oxides. The differences are in their colours, nodule morphology, frequency, and variety of bioclasts (Type 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiroCreatingFramesReference2023]. As seen in table @tbl-cortex_tab, whenever present, the cortex seems to show a mixture of irregular cortex and rounded cortex, both from an outcrop source. Petrographically, however, these cherts are homogeneous, except for a small number of artefacts with a banded variety within Type 5. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Type 5 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin section, they are sponge spicules (@fig-local-types b2, c2, h2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (@fig-local-types h3) in a sample from Type 5. Porosity varies between 10% and 1%, of vug type. Despite the petrographic homogeneity between local types, this macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data.

```{r}
#| label: tbl-cortex_tab
#| tbl-cap: Frequency of cortex type and thickness by raw material from the Terrace area, characterized during the individual analysis of the lithic assemblage.

@@

Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiroCreatingFramesReference2023], we interpret Types 2, 3, 4, and 5 to be local, as they are congruent with the described characteristics representative of cherts from the Lower Jurassic found in the known outcrops Cabo S. Vicente, Foz dos Fornos, Ponta dos Altos, Praia do Belixe and Ferrel (Carixian; \<20 km from Vale Boi, as the crow flies), possibly Upper Jurassic Kimmeridgian outcrops of Mareta (\~15 km from Vale Boi, as the crow flies) and Middle Jurassic Malhão formations (80-100 km East of Vale Boi, as the crow flies) of the Algarve basin. Types 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds (@fig-local-types b-d), ranging from uncommon to very frequent fossils. These can be found in all previously mentioned outcrops. The macroscopic similarities of Types 4 and 5 are more limited (@fig-local-types e-h), as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the inland Lower Jurassic outcrop of Ferrel (@fig-map b). As such, we interpret these cherts as local, but possibly from nodules and inland outcrops that are no longer frequently available or visible in the landscape. This is especially relevant since the inland outcrops of Ferrel are currently mostly destroyed and can be found in poor preservation conditions, due to recent human intevention and construction work, and with small, broken nodules in the surface, frequently altered [@belmiroCreatingFramesReference2023].

```{r}
#| label: fig-local-types
#| fig-cap: Variability of local chert types and main petrographic characteristics. a) Type 1 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope of white patina and fractures; a2) microscopic view of thin section (XPL), with visible fibrous chalcedony composition and without allochems. b) Type 3 macroscopic and microscopic views: b1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; b2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. c-d) Type 2 macroscopic and microscopic views of multicoloured (c) and single-colour (d) varieties. c1) macroscopic view with a stereomicroscope of the multicoloured bands with iron oxides and unidentifiable fossils; d1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils; c2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; d2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides. e-f) Type 4 macroscopic and microscopic views: e) macroscopic view with a stereomicroscope of unidentifiable fossils and white splotches; f1) macroscopic view with a stereomicroscope of iron oxides and unidentifiable fossils within white and yellow splotches; f2) microscopic view of thin section (XPL), with visible unidentifiable fossils and iron oxides; f3) microscopic view of thin section (PPL) with a concentration of fossils (possibly sponge spicules) and a fracture. g-h) Type 5 macroscopic and microscopic views: g) macroscopic view with a stereomicroscope; h1) macroscopic view with a stereomicroscope with visible unidentifiable fossils; h2) microscopic view of thin section (PPL), with visible unidentifiable fossils, sponge spicules and iron oxides; h3) microscopic view of thin section (XPL) with small unidentifiable fossils, iron oxides and a possible ostracod replaced by fibrous chalcedony and macrocrystalline quartz.

@@

### Cherts not congruent with local sources (Types 6-11 and TL)

@@

```{r}
#| label: fig-distance-types
#| fig-cap: Variability of non-local types and main petrographic characteristics. a-b) Type 6 macroscopic and microscopic views: a1) macroscopic view with a stereomicroscope with a small degree of white patina and no visible fossils; b1) macroscopic view with a stereomicroscope of a thick alteration rind but translucent interior where some unidentifiable fossils are visible; a2) microscopic view of thin section (PPL) with a mudstone texture; b2) microscopic of thin section (XPL) with small unidentifiable fossils and a laminar microstructure. c-d) Type 7 macroscopic and microscopic views of the peloidal facies (c) and oolitic facies (d): c1) macroscopic view with a stereomicroscope of peloids and sponge spicules; d1) macroscopic view with a stereomicroscope of poorly preserved ooids; c2) microscopic view of thin section (PPL) with packstone texture composed mainly of peloids; d2) microscopic view of thin section (PPL) with packstone texture composed mainly of ooids. e-f) Type 8 macroscopic and microscopic views, with foraminifera represented by the white and black arrows: e) macroscopic view with a stereomicroscope with abundant unidentifiable fossils and foraminifera; f1) macroscopic view with a stereomicroscope with abundant unidentifiable fossils but no visible foraminifera; f2) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which are not visible macroscopically; f3) microscopic view of thin section (PPL) with packstone texture composed of unidentifiable fossils and foraminifera which not are visible macroscopically. g) Type 9 macroscopic and microscopic views: g1) macroscopic view with a stereomicroscope of unidentifiable fossils and two different fabrics; g2) microscopic view of thin section (XPL) with packstone texture composed of concentrations of unidentifiable fossils and opaques which create a banded microstructure. h) Type 10 macroscopic and microscopic views: h1) macroscopic view with a stereomicroscope of ooids and fractures filled with chalcedony; h2) microscopic view of thin section (PPL) with packstone texture composed of ooids with different levels of preservation.

@@

Studies of Upper Cretaceous cherts located in the Lisbon area describe cherts with similar macroscopic characteristics (@fig-map a), with micro-cryptocrystalline quartz mudstones with bioclasts such as sponge spicules or bivalve fragments, or wackestone/packstone textures with abundant bioclasts, including ostracods and different types of foraminifers [@jordaoFlintSourcesMobility2022]. These cherts have also been reported as found in highly variable cobble morphologies and macroscopic traits, which is also observable through the reference samples from Central Portugal hosted at LusoLit. Alike the archaeological thin sections from Type 6 and congruent with other studies [@jordaoFlintSourcesMobility2022; @matiasRawMaterialSourcing2016], the thin sections produced from the Cretaceous cherts hosted at Lusolit show a massive, cryptocrystalline mudstone structure, with uncommon fossils (ghosts), commonly barely visible, iron oxides and low porosity (\<1%). From these Cretaceous cherts from Central Portugal, only the dark grey nodules (sample RT231, with petrographic descriptions available in our online research compendium) showed identifiable fossils which were rare ostracods.

@@
